---
title: "Assignment 2: Coding Basics"
author: "Kendall Barton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1.
sequence3 <- seq(1,30,3) #number sequence 1-30 by 3s
sequence3
#2. 
#find mean and median and assign names
s3mean <- mean(sequence3)
s3median <- median(sequence3)
#3. 
s3mean > s3median #use names to compare mean and median
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5 & 6
students <- c("Matt","Steve","Kevin","Priscilla") #name vector
test_scores <- sample(0:100,4) #4 random test scores 0-100
print(test_scores)
passing <- test_scores >= 50 #pass/fail vector
print(passing)
#7 & 8 
test_df <- data.frame("Student Name" = students, "Test Score" = test_scores, "Pass/Fail" = passing)
print(test_df)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frames can hold different types of data, while all data in a matrix has to be the same kind.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#I feel like this is easier to do without a function
pass_function<- function(scores){
  for(score in scores){
    if(score>=50) {print(TRUE)}
    else {print(FALSE)}}}

pass_function(test_scores)

pass_function2 <- function(scores){
    print(ifelse(scores>=50, TRUE, FALSE))
  }

pass_function2(test_scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Both options work equally. Because there are only two options (TRUE and FALSE), ifelse can be used
