---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Kendall Barton"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(lubridate)
library(ggthemes)

getwd()

ntl_chemphys_df <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
                        stringsAsFactors = TRUE)
ntl_chemphys_df$sampledate <- mdy(ntl_chemphys_df$sampledate) #using lubridate to class dates

#2
my_theme <- theme_base() +
  theme(
    plot.title = element_text(
      hjust = 0, size = 16, face = "bold"
    ),
    panel.grid = element_line( #add grid
      color = "gray"
    ),
    axis.ticks = element_blank( #no axis ticks
    ),
    legend.text = element_text(
      size = 10
    ),
    axis.title = element_text( #x and y labels
      size = 12
    ),
    axis.text = element_text( #numbers
      size = 10
  ))
theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature during July does not change with depth across all lakes.
Ha: Mean lake temperature during July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
wrang_ntl_df <- ntl_chemphys_df %>%
  filter(month(sampledate) == 7) %>% #only July
  select(lakename:temperature_C, -sampledate) %>% #only certain columns
  na.omit() #no nas

#5
ggplot(wrang_ntl_df, aes(y = temperature_C, x = depth)) +
  geom_point() +
  labs(title = "July Lake Temperature by Depth", y = "Temperature (C)", x = "Depth (m)") +
  geom_smooth(method = "lm") + #linear model
  ylim(0, 35) #temp only 0-35

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure suggests that temperature does decrease with depth down to about 7 meters before becoming more constant.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp_reg <- lm(data = wrang_ntl_df, temperature_C ~ depth)
print(summary(temp_reg))
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 73.87% of the variance in the temperature data is explained by depth. There are 9726 degrees of freedom and a p-value of of less than 2.2e-16. The p-value only has to be less than 0.05 to be significant, so this result is defintely significant. There is a predicted change of -1.94621 degrees C with every meter deeper change.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp_aic <- lm(data = wrang_ntl_df, temperature_C ~ depth + year4 + daynum)
print(step(temp_aic))
#all three should improve model at least somewhat

#10
temp_mult_reg <- lm(data = wrang_ntl_df, temperature_C ~ depth + daynum + year4)
print(summary(temp_mult_reg))
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: It suggests that we use depth, daynum and year4 to predict temperature, but daynum and especially year4 don't explain much of the variance. Our new model explains 74.11% of the variance, which is less than a 1% increase over the previous model.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
ntl.anova <- aov(data = wrang_ntl_df, temperature_C ~ lakename)
summary(ntl.anova)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in temperature among the lakes because the p-value is <2e-16, which is certainly less than 0.05.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(wrang_ntl_df, aes(y = temperature_C, x = depth, color = lakename)) +
  geom_point(alpha = 0.5) + #transparency
  labs(title = "July Lake Temperature by Depth in Lakes", y = "Temperature (C)",
       x = "Depth (m)", color = "") + #no legend title, it seems self-explanatory
  geom_smooth(method = "lm") + #linear model
  ylim(0, 35) #temp only 0-35
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(ntl.anova)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: According to the Tukey test, Paul Lake and Ward Lake have the same mean temperature, statisitically speaking, as Peter Lake with p-values of 0.2241586 and 0.7827037 respectively.

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could run a two-sample t-test to compare the mean temperatures of just the two lakes.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
cr_ward_data <- wrang_ntl_df %>%
  filter(lakename == "Crampton Lake"| lakename == "Ward Lake") #only Ward or Crampton lake
t.test(cr_ward_data$temperature_C ~ cr_ward_data$lakename)
```

>Answer: The p-value is too large to be significant (0.2649), and the 95% confidence interval includes 0, so the mean temperatures of the Crampton Lake and Ward Lake are not significantly different in July. This matches what I saw in the Tukey test, where the p-value was 0.9714459.

