---
title: "Assignment 5: Data Visualization"
author: "Kendall Barton"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder).

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggthemes)
getwd()
lake_chem_df <- read.csv("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
litter_df <- read.csv("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2 
class(lake_chem_df$sampledate)
class(litter_df$collectDate)
lake_chem_df$sampledate <- as.Date(lake_chem_df$sampledate)
litter_df$collectDate <-as.Date(litter_df$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
theme1 <- theme_base() + 
  theme(
    plot.title = element_text(
      color = 'gray35', hjust = 0.5, size = 16
    ),
      plot.background =   element_rect(
      fill = 'mistyrose', color = 'gray35'
    ),
    legend.background = element_rect(
      color='gray35',
      fill = 'rosybrown'
    ),
    legend.title = element_text(
      color='white', size = 12
    ),
    legend.text = element_text(
      size=10),
    axis.title = element_text(
      size=12),
    axis.text = element_text(
      size=10)
  )
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
# x = po4, y = tp_ug
ggplot(lake_chem_df, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +
  xlim(0,35) +
  labs(title = "Phosphorus by Phosphate in Peter and Paul Lakes", x = "Phosphate", 
       y = "Phosphorus", color = "Lake Name") +
  geom_smooth(method = lm, color = "black") +
  theme1
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
temp_bp <- ggplot(lake_chem_df, 
                  aes(x = factor(month, levels = 1:12, 
                                 labels = month.abb), 
                      y = temperature_C, color = lakename)) +
  geom_boxplot() +
  scale_x_discrete(name = "Month", drop = F) +
  labs(title = "Lake Temperature by Month", y = "Temperature (C)",
       color = "Lake Name") +
  theme1
temp_bp

tp_bp <- ggplot(lake_chem_df, 
                aes(x = factor(month, levels = 1:12,
                               labels = month.abb), 
                    y = tp_ug, color = lakename)) +
  geom_boxplot() +
  scale_x_discrete(name = "Month", drop = F) +
  labs(title = "Lake Total Phosphorus by Month", x = "Month", y = "Total Phosphorus",
       color = "Lake Name") +
  theme1
tp_bp

tn_bp <- ggplot(lake_chem_df,
                aes(x = factor(month, levels = 1:12,
                                             labels = month.abb),
                                  y = tn_ug, color = lakename)) +
  geom_boxplot() +
  scale_x_discrete(name = "Month", drop = F) +
  labs(title = "Lake Total Nitrogen by Month", x = "Month", y = "Total Nitrogen",
       color = "Lake Name") +
  theme1
tn_bp

plot_grid(temp_bp + theme(legend.position="none"),
          tp_bp + theme(legend.position="none"),
          tn_bp + theme(legend.position="none"), ncol = 1, align = "h", axis = "lr",
          legend = get_legend(temp_bp + 
                                guides(color = guide_legend(nrow = 1)))) +
  theme1


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
ggplot(filter(litter_df, functionalGroup == "Needles"),
       aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  labs(title = "Needle Dry Mass by Date and NCLD Class", x = "Sample Date",
       y = "Needle Dry Mass", color = "NLCD Class") +
  theme1
  
#7
ggplot(filter(litter_df, functionalGroup == "Needles"),
       aes(x = collectDate, y = dryMass)) +
  geom_point() +
  labs(title = "Needle Dry Mass by Date and NCLD Class", x = "Sample Date",
       y = "Needle Dry Mass (g)") +
  facet_wrap(vars(nlcdClass)) +
  theme1

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I like the faceted graph (7) better. The points in 6 are somewhat overlapping, which makes reading the graph hard. Additionally, I think it would be easier for people with colorblindness to read graph 7 because color isn't as important of a component.

